---
title: "Week 8: Strings, dates and times, joining and binding"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
```

class: inverse, center, middle

# Strings, dates and times, joining and binding

---

# Data wrangling ?

- We've discussed select(), filter(), group(), mutate() etc.
- A lot of data wrangling is about fixing strings, dates, joining different sources of data

---

# Plan for today

- Joining and binding
- Dealing with character strings
- Dealing with dates

# Binding (columsn or rows) means appending new lines or new columns

## Binding rows

- `bind_rows()` is often used if you want to stack some data together than has the same variables

```{r }

df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
df2 <- tibble(age=c(55,30,12),name=c("Hillary","Anna","Luis"))
(df <- bind_rows(df1,df2))
```

---

```{r eval=FALSE}
df1 <- read_csv("mydata1.csv")
df2 <- read_csv("mydata2.csv")
df <- bind_rows(df1,df2)
```

- Sometimes, you can use a loop to read in numerous csvs. We'll cover this next week. Think of data for 125 riding.
- Need to make sure columns names are the same.

--- 

# Binding columns

- `bind_cols()` is often used if you want to stack variables where the individuals are the same

```{r }

df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
df2 <- tibble(income=c(55,30,12),name=c("Justin","Ian","Rosaline"))
(df <- bind_cols(df1,df2 %>% select(-name)))
```

---

```{r }

(df <- bind_rows(df1,df2 %>% select(-name)))
```

---

# Joins

- `inner_join()`: includes all rows in x and y.

- `left_join()`: includes all rows in x.

- `right_join()`: includes all rows in y.

- `full_join()`: includes all rows in x or y.

---

```{r}
left_join(df1,df2)
#left_join(df1,df2,c("name"))
#left_join(df1,df2,c("name"="name"))
```

---

```{r}
df1 <- tibble(age=c(25,29,22,23),name=c("Justin","Ian","Rosaline","Jack"))
df2 <- tibble(income=c(55,30,12),name=c("Justin","Ian","Rosaline"))
inner_join(df1,df2)
left_join(df1,df2)
```

---

```{r}
left_join(df2,df1)
```

---

```{r}
df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
df2 <- tibble(income=c(55,30,12,33),name=c("Justin","Ian","Rosaline","Justin"))
left_join(df1,df2)
```

---

```{r}
df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"),last=c("Savoie","Smith","Azevedo"))
df2 <- tibble(income=c(55,30,12,33),name=c("Justin","Ian","Rosaline","Justin"),last=c("Savoie","Smith","Azevedo","Gajevic"))
left_join(df1,df2)
```

--- 

# Dealing with character strings

```{r}
my_string <- c("Justin Savoie (2019)","Daniel Ortega (2019)", "Natalia Burina (2019)")
str_replace(my_string,"\\(2019\\)","") %>% trimws()
my_string <- c("POL2517-U","POL2517-V","POL2699-S")
str_sub(my_string,4,-1)
str_sub(my_string,4,7)
my_string <- c("Justin Savoie (2019)","Daniel Ortega (2017)", "Natalia Burina (2018)")
str_extract(my_string,"(?<=\\().+?(?=\\))")
str_replace_all(my_string,"\\((?<=\\().+?(?=\\))\\)","") %>% trimws()
# https://stackoverflow.com/questions/8613237/extract-info-inside-all-parenthesis-in-r
```

---

```{r}
my_string <- c("justin savoie","micheal Gillet","daniela Smith-jones")
str_to_title(my_string)
str_length(my_string)
str_count(my_string,"j")
```

---

# Dates and times

```{r}
library(lubridate)
today()
now()+60
today()+1
```

---

```{r}
adate <- "2022-06-01"
ymd(adate)
lubridate::dmy("26-06-19")
adatetime <- "2001-10-10 02:30:30"
ymd_hms(adatetime)

as.numeric(ymd_hms(adatetime))

```

---

```{r}
to_plot <- tibble(date=c("2019-01-01","2019-03-01","2020-01-01","2020-06-01"),value=rnorm(4,10,3))
ggplot(to_plot,aes(x=date,y=value)) + geom_point()
```

```{r}
to_plot <- to_plot %>% mutate(date=ymd(date))
ggplot(to_plot,aes(x=date,y=value)) + geom_point() + geom_line() +
  scale_x_date(breaks = as_date(c("2019-01-01","2019-03-01","2020-01-01","2020-06-01")),
               labels=c("Jan 2019","March 2019","Jan 2020","June 202")) +
  theme(axis.text.x = element_text(angle=45)) + scale_y_continuous(limits=c(0,30))
```

---

# An example

```{r eval=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)

df <- read_excel("~/Downloads/Parliamentarians.xlsx")
profession <- read_excel("~/Downloads/Parliamentarians Job.xlsx")

df <- left_join(df,profession)

head(df$`Province/Territory`)

df <- df %>%
  mutate(province=str_extract(`Province/Territory`, "[^\r]+"))

head(df$`Type of Parliamentarian`)

df <- df %>%
  mutate(start_date=str_extract(`Type of Parliamentarian`, "\\([^()]+\\)") %>%
           str_sub(2,11) %>%
           ymd())

df %>%
  select(Name,start_date,Profession,province)
```

---

## Two questions from the Project

```{r eval=FALSE}
hsum <- df %>%
  summarize(mean_age = mean(hincome, na.rm = TRUE), 
            sd_age = sd(hincome, na.rm = TRUE), 
            min_age = min(hincome, na.rm = TRUE), 
            max_age = max(hincome, na.rm = TRUE), 
            median_age = median(hincome, na.rm = TRUE), 
            skew_age = skewness(hincome, na.rm = TRUE), 
            kurtosis_age = kurtosis(hincome, na.rm = TRUE), 
            n_age =  sum(!is.na(hincome)))

hsum %>%
  kable(format = "simple") 

hsum %>%
  t() %>%
  kable(format = "simple") 

hsum %>%
  pivot_longer(cols = everything()) %>%
  mutate(value=as.character(round(value))) %>%
  kable(format = "simple") 
```
---
```{r eval=FALSE}
kurtosis(df$hincome,na.rm = TRUE)

options(scipen=999)
unname(quantile(df$hincome,0.99,na.rm=TRUE))
temp <- df$hincome[df$hincome <=500000]

kurtosis(temp,na.rm = TRUE)
temp <- df$hincome
kurtosis(temp,na.rm = TRUE)


```
---
# A worked example instead of 