---
title: "Week 3: Data transformation"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(kableExtra)
library(tidyverse)
```

class: inverse, center, middle

# Summary Week 2 Data Viz

---

```{r}
ggplot(mpg) +
  geom_point(aes(x=displ,y=hwy, color=class))
```

---

```{r}
ggplot(mpg) +
  geom_point(aes(x=displ,y=hwy, color=factor(year))) +
  facet_wrap(~class) +
  coord_polar()
```

---

```{r message=FALSE, fig.align='center'}
library(maps)
world <- map_data("world")
world$type <- "No"
world$type <- ifelse(world$region %in% c("China","France","Russia","UK","USA"),"Perm.",world$type)
world$type <- ifelse(world$region %in% c("Albania","Brazil","Gabon","Ghana","India","Ireland","Kenya","Mexico","Norway","United Arab Emirates"),"Non.perm",world$type)

ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group, fill=type)) + 
  labs(fill="UN Security Council") + coord_fixed(1.3) + theme(legend.position = "top", plot.margin = margin(-5,1,1.5,1.2, "cm"))
```

---
class: inverse, center, middle

# Basics

---

### Objects

```{r}
x <- c(1,2,3)
x
```

```{r}
x <- "my name is justin"
x
```

```{r}
z <- sum(c(1,2,3))*2
z
```

```{r}
x1 <- c(2,4,6)
x2 <- c(4,5,1)
x12 <- x1+x2
x12
```

---

### Functions

```{r}
my_vector <- c(1,2,3,4,5)
mean(my_vector)
```

```{r}
rnorm(10,mean=100,sd=25)
```

```{r}
seq(from=0,to=10,by=2)
```
```{r}
seq(0,10,2)
```
---
class: inverse, center, middle

# Data transformation

---

### Five key verbs

`filter()`
`arrange()`
`select()`
`mutate()`
`summarise()`

can be used with 
`groub_by()`

which applies the verbs, but "by group"

---

### Let's look at the flights data frame

```{r}
library(nycflights13)
flights
```

---

### filter()

```{r}
filter(flights,month==1,day==1)
```

### filter()

```{r}
filtered_flights <- filter(flights,month==1,day==1)
filtered_flights
```

### filter()

```{r}
filter(flights,arr_time>2330)
```
---

### filter()

```{r}
filter(flights,arr_time %in% c(2200,2300,2400))
```
---

### filter()

```{r}
filter(flights,is.na(arr_time))
```
---

### arrange()

```{r}
head(arrange(flights,arr_time))
```

```{r}
head(arrange(flights,desc(arr_time)))
```

```{r}
head(arrange(flights,month,desc(day)))
```

---

### select()
```{r}
select(flights,month,day)  
```
---

### select()
```{r}
select(flights,-month,-day)  
```
---

### select()
```{r}
select(flights,contains('time'))
```
---

### mutate()
```{r}
mutate(mtcars,one=1)
```
---

### mutate()
```{r}
mutate(mtcars,one=1)
```
---
### mutate()
```{r}
mutate(mtcars,hp2=hp^2)
```
---
### mutate()
```{r}
mutate(mtcars,hpbycyl=hp/cyl)
```
---
### group_by() with mutate()
```{r}
mtcars_grouped <- group_by(mtcars,cyl)
mutate(mtcars_grouped,avg_mpg_cyl=mean(mpg))
```
---
### group_by() with summarise()
```{r}
mtcars_grouped <- group_by(mtcars,cyl)
summarise(mtcars_grouped,avg_mpg_cyl=mean(mpg))
```
