---
title: "Week 3: Data transformation"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(kableExtra)
library(tidyverse)
```

class: inverse, center, middle

# Summary Week 2 Data Viz

---

### Basic visualization using ggplot2

```{r, eval=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
# initializes a ggplot object
ggplot(penguins,mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
  # point geom is used to create scatterplot
  geom_point(aes(color = species, shape = species)) +
  # add a visual line of best fit
  geom_smooth(method = "lm") +
  # add labels
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) + 
  # change the color palette
  scale_color_colorblind()
```

---

```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
ggplot(penguins,mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) + scale_color_colorblind()
```

---
class: inverse, center, middle

# Coding basics

---

### Objects

```{r}
x <- c(1,2,3)
x
```

```{r}
x <- "my name is justin"
x
```

```{r}
z <- sum(c(1,2,3))*2
z
```

```{r}
x1 <- c(2,4,6)
x2 <- c(4,5,1)
x12 <- x1+x2
x12
```

---

### Assigning a value the value of something to an object

```{r, eval=FALSE}
object_name <- value
```

```{r, eval=FALSE}
x1 <- c(2,4,6)
print(x1)
x1
```

---

### Comments

```{r}
# you can leave comments using the like this using the #
# create vector of primes
primes <- c(2, 3, 5, 7, 11, 13)
```

**Great advice:** Use comments to explain the why of your code, not the how or the what. Why did you use this model instead of that model. Why did you use binwidth of 2.3 instead of 2.8. If when you code you make a decision that is not fully trivial, it makes sense to write a comment.

In general, if you start using R more frequently, you should try to make things clear and replicable. There's nothing as frustrating as going back in old code when you need it and not understanding or not remebering why you did something.

---

### naming objects

You can give different names to objects

```{r}
xYUBVChh.1 <- c(1,2,3)
xyubvchh_1 <- c(1,2,3)
```

The recommendation is to use snake_case where you use lowercase words with `_`.

These are ok:

```{r}
x <- c(1,2,3)
my_vector <- c(1,2,3)
```

---

### Functions

```{r}
my_vector <- c(1,2,3,4,5)
mean(my_vector)
```

```{r}
rnorm(10,mean=100,sd=25)
```

```{r}
seq(from=0,to=10,by=2)
```
```{r}
seq(0,10,2)
```
---
class: inverse, center, middle

# Data transformation

---

### Five key verbs (or functions) in dplyr

- It's very rare we get the data in the form we need it. We need to transform it first.

- dplyr has verbs: verbs are just functions that do things

- These verbs do things on **rows** or on **columns** or on **groups** or on **tables** (or data frames). This week we talk about rows, columns and groups.

- The pipe (it is not a verb) "`|>`" (it used to be `%>%` so you will see that often), but use "`|>`" allows to combine verbs.

---

### verbs on rows

`filter()` (keep some rows)

`arrange()` (reorder the rows)

`distinct()` (keep only unique rows)

---

### Let's look at the flights data frame

```{r}
library(nycflights13)
flights
```

---

### filter()

```{r}
filter(flights,month==1,day==1)
```
---
### filter()

```{r}
flights |> 
  filter(dep_delay > 120)
```
---
### filter()

```{r}
flights |> 
  filter(month == 1 & day == 1)
```
---

### filter()

```{r}
flights |> 
  filter(month == 1 | month == 2)
```
---

### filter()

```{r}
flights |> 
  filter(month %in% c(1, 2))
```
---

### filter()

```{r eval=FALSE}

flights |> 
  filter(month == 1 & day == 1)

jan1 <- flights |> 
  filter(month == 1 & day == 1)

```

---

### common mistake

```{r eval=FALSE}
flights |> 
  filter(month = 1)
flights |> 
  filter(month == 1)
```

---

### arrange()

```{r}
flights |> 
  arrange(year, month, day, dep_time)
```

---

```{r}
flights |> 
  arrange(desc(dep_delay))
```
---

### distinct()

```{r}
flights |> 
  distinct()
```

---

### distinct()

```{r}
flights |> 
  distinct(origin, dest)
```

---

### count()

```{r}
flights |>
  count(origin, dest, sort = TRUE)
```

---


### verbs on columns

`mutate()` (creates a new variable)

`select()` (keeps only some variables)

`rename()` (renames variables)

`relocate()` (changes the positions of variables)

---

### mutate()

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```

---

### mutate()

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )
```

---

### select()
```{r}
flights |> 
  select(year, month, day)
```
---

### select()
```{r}
flights |> 
  select(year:day)
```
---

### select()
```{r}
flights |> 
  select(!year:day)
```
---

### select()

```{r}
flights |> 
  select(where(is.character))
```

---

### rename()
```{r}
flights |> 
  rename(tail_num = tailnum)
```

---

### relocate()

```{r}
flights |> 
  relocate(time_hour, air_time)
```
---

### relocate()

```{r}
flights |> 
  relocate(year:dep_time, .after = time_hour)
```

---

### the pipe `|>` 

Allows to combine multiple verbs.

```{r}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```
---

```{r ,eval=FALSE}
arrange(
  select(
    mutate(
      filter(
        flights, 
        dest == "IAH"
      ),
      speed = distance / air_time * 60
    ),
    year:day, dep_time, carrier, flight, speed
  ),
  desc(speed)
)

flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```
---

### working with groups with group_by()

```{r}
flights |> 
  group_by(month)
```

---

### group_by() and summarize()

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE)
  )
```

---

### group_by() and summarize()

.pull-left[

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay)
  )
```
]
.pull-right[

```{r  out.width='50%',warning=FALSE}
x <- c(1,2,3,NA)
mean(x)
mean(x,na.rm=TRUE)
```
]

---

### group_by() and summarize()

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )
```

---

### group_by() and summarize()

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )
```

---

### `slice_`

```{r eval=FALSE}
df |> slice_head(n = 1) # first
df |> slice_min(x, n = 1) # smallest value
```

```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

---
class: inverse, center, middle

# Homework 1

---

class: inverse, center, middle

# Exercices

### 4.2.5, 4.3.5, 4.5.7