---
title: "Week 4: Importing data and tidy data"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(kableExtra)
library(tidyverse)
```

class: inverse, center, middle

# Homework 1
### Solutions code posted on the website
---

class: inverse, center, middle

# The pipe (%>%)
---

# The pipe (%>%)

- We use the pipe to emphasise a sequence of actions
- The pipe means "THEN"

```{r}
mtcars_grouped <- group_by(mtcars,cyl)
count(mtcars_grouped)
```
---
# The pipe (%>%)
```{r}
mtcars %>%
  group_by(cyl) %>%
  count()
```

In plain language: take the mtcars dataset THEN group by "cyl" THEN count.
---

```{r}
mtcars %>%
  filter(cyl==4) %>%
  select(mpg,hp,vs,cyl)
```
---
# Another way to think of the pipe (%>%)
```{r}
my_vector <- c(1,2,3,4,8)
mean(my_vector)
my_vector %>%
  mean()
```

More formally, the pipe passes what comes before the pipe as the first argument in the function after the pipe.
---
# Now that we know about the pipe, let's take a look at the homework
---

class: inverse, center, middle

# Data importing
---

# Reading a csv

```{r, eval=FALSE}
library(tidyverse)
Parliamentarians_subset100 <- read_csv("https://www.justinsavoie.com/pubpol750-2022/homework1/Parliamentarians_subset100.csv")
Parliamentarians_subset100 <- read_csv("~/Downloads/Parliamentarians_subset100.csv")
Parliamentarians_subset100 <- read_csv("/Users/justinsavoie/Downloads/Parliamentarians_subset100.csv")
Parliamentarians_subset100 <- read_csv("/Users/justinsavoie/Documents/personal_repos/personal_website/pubpol750-2022/Homework1/Parliamentarians_subset100.csv")
Parliamentarians_subset100 <- read_csv("C:\\Users\\justinsavoie\\Downloads/Parliamentarians_subset100.csv")
```

# Reading an excel file

```{r}
library(readxl)
Parliamentarians_subset100 <- read_excel("~/Downloads/Parliamentarians.xlsx")
```
---
# Reading spss, sas, stata files
```{r}
library(haven)
#read_spss()
#read_sas()
#read_stata()
library(readstata13)
#readstata13::read.dta13()
```

---

class: inverse, center, middle

# Tidy data
---

# Tidy data

(A)
- Each variable must have its own column.
- Each observation must have its own row.
- Each value must have its own cell.

(B)
- Put each dataset in a tibble.
- Put each variable in a column.

---

.pull-left[
```{r}
table1
table2
```
]
.pull-right[
```{r}
table3
table4a
table4b
```
]
---
# Pivoting

- Pivoting wider
- Pivoting longer
---
# Pivoting wider

- column names are not names of variables, but values of a variable
.pull-left[
```{r}
table4a
```
]
.pull-right[
```{r}
table4a %>%
  pivot_longer(cols=c(`1999`,`2000`),names_to = 'year', values_to = 'cases')
```
]
---
```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```
---
# Pivoting wider
.pull-left[
```{r}
table2
```
]
.pull-right[
```{r}
table2 %>%
    pivot_wider(names_from = type, values_from = count)
```
]
---
# Separating
.pull-left[
```{r}
table3
```
]
.pull-right[
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```
]
---
# Uniting
.pull-left[
```{r}
table5
```
]
.pull-right[
```{r}
table5 %>% 
  unite(new, century, year)
```
]
---
# Missing values

- Explicit vs Implicit

```{r}
(stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
))
```

---

- Force explicit

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```

---

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>%
  pivot_longer(cols=c(`2015`,`2016`),names_to='year',values_to='return')
```
---

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>%
  pivot_longer(cols=c(`2015`,`2016`),names_to='year',values_to='return',values_drop_na = TRUE)
```
---
.pull-left[
```{r}
stocks
```
]
.pull-left[
```{r}
stocks %>% 
  complete(year, qtr)
```
]