---
title: "Week 9: Functions, loops"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
```

class: inverse, center, middle

# Becoming more efficient: functions and loop

---

# Becoming more efficient with functions and loop

- Automate common tasks
- Apply the same "procedure" to tens or hundreds of files, of objects, of dataframes, of variables.


- In a nutshell: 
  - Functions allow to avoid copy pasting, allow to reuse a procedure
  - Loops allow to do one thing many times
  
They are often used in combination.

---

## Functions

```{r}
set.seed(232)
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$b, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
head(df)
```
---
```{r}
set.seed(232)
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10)
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
head(df)
```
---

Implementing the mean function by taking the sum, and dividing by the number of non missing cases.

```{r}
my_vector <- c(3,8,19,NA,22,1)
mean(my_vector,na.rm=TRUE)
custom_mean1 <- function(x){
  my_sum <- sum(x,na.rm=TRUE)
  my_length <- sum(!is.na(x))
  my_sum / my_length
}
custom_mean1(my_vector)
```
---

Implementing the same mean function by going through all values one by one and calculating a bit like we would do in elementary school. This is often the logic in real functions. Breaking down everything.

```{r}
custom_mean2 <- function(x){
  my_length <- 0
  my_sum <- 0
  for (i in my_vector){
    if (!is.na(i)){
      my_length <- my_length + 1
      my_sum <- my_sum + i
    }
  }
  my_sum / my_length
}
custom_mean2(my_vector)
```
---
## Existing functions

Of course, you likely won't have to write many functions. A lot of functions exists. We've been using functions since the beginning of the semester.

`mean()` `group_by()` `ggplot()` (and many more) these are all functions.

Functions are incredibly useful.

You might, for example, want to:

- Calculate cross tabs and plot them, for many survey questions (à la what polling firms)
- Read in a court decision and clean it to a format you can analyze

---

## Example 

Imagine we have 200 respondents on 10 questions. Answers: {1:"Strongly disagree",2:"Somewhat disagree",3:"Neutral",4:"Somewhat agree",5:"Strongly agree",6:"Don't know"}. We want to plot the response to each question in a separate figure.

```{r, warning=FALSE,message=FALSE}
questions <- paste0("question",1:10)
df <- as_tibble(matrix(sample(1:6,size=2000,replace = TRUE),nrow=200))
names(df) <- questions
df
```
---
```{r}

make_plot <- function(data,my_variable,answer_choices){
  
  to_plot <- data %>%
    select(all_of(my_variable)) %>%
    rename(selected_variable=all_of(my_variable)) %>%
    group_by(selected_variable) %>%
    count() %>%
    ungroup() %>%
    mutate(p=n/sum(n))  

  to_plot <- to_plot %>%
    mutate(selected_variable=recode(selected_variable,!!!answer_choices)) %>%
    mutate(selected_variable = factor(selected_variable,answer_choices))
  
  ggplot(to_plot,aes(x=selected_variable,y=p)) +
    geom_bar(stat='identity') +
    geom_text(aes(label = p), vjust = 1.5, colour = "white")
  
}
```
---
```{r}
answer_choices=c("Strongly disagree","Somewhat disagree","Neutral",'Somewhat agree',"Strongly agree","Don't know")
make_plot(data=df,my_variable='question1',answer_choices=answer_choices)
```
---
```{r}
make_plot(data=df,my_variable='question2',answer_choices)
```
---
```{r}
make_data_to_plot <- function(data,my_variable){
  
  to_plot <- data %>%
    select(all_of(my_variable)) %>%
    rename(selected_variable=all_of(my_variable)) %>%
    group_by(selected_variable) %>%
    count() %>%
    ungroup() %>%
    mutate(p=n/sum(n))  
  to_plot <- to_plot %>%
    mutate(selected_variable=recode(selected_variable,!!!answer_choices)) %>%
    mutate(selected_variable = factor(selected_variable,answer_choices))
  to_plot
}
make_plot <-  function(data,answer_choices){
  ggplot(data,aes(x=selected_variable,y=p)) +
    geom_bar(stat='identity') +
    geom_text(aes(label = p), vjust = 1.5, colour = "white")
}
```

---

```{r}
data_to_plot <- make_data_to_plot(data=df,my_variable='question2')
make_plot(data_to_plot,answer_choices)
```

---

## Loops (Iteration chapter in the book)

```{r}

threeletters <- c("a","b","c")

for (i in threeletters){
  print(i)
}
```
---

```{r}
for (i in seq_along(threeletters)){
  print(i)
  print(threeletters[i])
}

```

---

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

```{r}
output <- vector("double", ncol(df))
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```
---
```{r}
for (i in seq_along(df)) {            # 2. sequence
  df[[i]] <- df[[i]]+10      # 3. body
}
df
```

---

class: inverse, center, middle

# Exercices

### 19.2.1, 21.2.1



