---
title: "Week 8: Transforming data: more tricks"
subtitle: "PUBPOL 750 Data Analysis for Public Policy I"
author: "Justin Savoie"
institute: "MPP-DS McMaster"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
```

class: inverse, center, middle

# Administrative note
### Linear regression next week (Week 9). Functions and loops in Week 11.   

---

# Data wrangling ?

- `dplyr` verbs (e.g., select(), filter(), group(), mutate() etc.) perform specific tasks on data frames

- Data wrangling can also involve fixing messy strings, dates; joining data from different sources

---

# Plan for today

- Binding data frames

- Mutating joins

- Dealing with character strings

- Dealing with dates

- Worked example

---

# Binding (columsn or rows) means appending new lines or new columns

## Binding rows with `bind_rows()`

- Bind any number of data frames by row, making a longer result.

```{r }

df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
df2 <- tibble(age=c(55,30,12),name=c("Hillary","Anna","Luis"))
(df <- bind_rows(df1,df2))
```

---

```{r eval=FALSE}
df1 <- read_csv("mydata1.csv")
df2 <- read_csv("mydata2.csv")
df <- bind_rows(df1,df2)
```

- Sometimes, you can use a loop to read in numerous csvs. We'll cover this in week 11. Think of data for 338 ridings.

- For `bind_rows()` columns names have to be the same; or if they are not it's simply to create a new column.

```{r}
df1 <- tibble(age1=c(25,29,22),name1=c("Justin","Ian","Rosaline"))
df2 <- tibble(age2=c(55,30,12),name2=c("Hillary","Anna","Luis"))
(df <- bind_rows(df1,df2))
```

- In this case you would have to rename the columns.

---

## Binding columns with `bind_cols()`

- Bind any number of data frames by column, making a wider result.

```{r}

df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
df2 <- tibble(income=c(55,30,12),name=c("Justin","Ian","Rosaline"))
(df <- bind_cols(df1,df2 |> select(-name)))
```

---

```{r}
(df <- bind_cols(df1,df2))
```

---

# Joins

Instead of "stacking", with a bind, joins "joins" on one or more columns. If you have riding names in two different datasets, you can join them.

- `inner_join()`: includes all rows in x and y.

- `left_join()`: includes all rows in x.

- `right_join()`: includes all rows in y.

- `full_join()`: includes all rows in x or y.

FYI: the notion of 'joins' are a fundamental aspect of SQL (Structured Query Language, is a standardized programming language that is used to manage and manipulate relational databases; SQL is a dominant language for interacting with databases) and are used extensively. They allow you to combine rows from two or more tables based on a related column between them, enabling the construction of complex queries over relational data.

---

```{r}
df1;df2
left_join(df1,df2,by="name")
```

---

```{r}
left_join(df1,df2,by="name")
left_join(df1,df2)
```

---

```{r}
df1 <- tibble(age=c(25,29,22,23),name=c("Justin","Ian","Rosaline","Jack"))
df2 <- tibble(income=c(55,30,12),name=c("Justin","Ian","Rosaline"))
left_join(df1,df2,by='name')
```

---

```{r}
inner_join(df1,df2,by='name')
```

---

```{r}
left_join(df1,df2,by='name')
right_join(df2,df1,by='name')
```

---

```{r}
full_join(df1,df2,by='name')
```

---

```{r}
df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"))
(df2 <- tibble(income=c(55,30,12,33),name=c("Justin","Ian","Rosaline","Justin")))
left_join(df1,df2,by='name')
```

---

```{r}
df1 <- tibble(age=c(25,29,22),name=c("Justin","Ian","Rosaline"),last=c("Savoie","Smith","Azevedo"))
df2 <- tibble(income=c(55,30,12,33),name=c("Justin","Ian","Rosaline","Justin"),last=c("Savoie","Smith","Azevedo","Gajevic"))
left_join(df1,df2,c("name"="name","last"="last"))
```

---

# Dealing with character strings

- Very useful when dealing with text

- Cryptic at first



- replacing, substituting, extracting with `stringr` package

---

```{r}
my_string <- c("Justin Savoie (2019)","Daniel Ortega (2019)", "Natalia Burina (2019)")
str_replace(my_string,"\\(2019\\)","") %>% trimws()
my_string <- c("POL2517-U","POL2517-V","POL2699-S")
str_sub(my_string,4,-1)
str_sub(my_string,4,7)
my_string <- c("Justin Savoie (2019)","Daniel Ortega (2017)", "Natalia Burina (2018)")
str_extract(my_string,"(?<=\\().+?(?=\\))")
str_replace(my_string,"\\((?<=\\().+?(?=\\))\\)","") %>% trimws()
# https://stackoverflow.com/questions/8613237/extract-info-inside-all-parenthesis-in-r
```

---

```{r}
my_string <- c("justin savoie","micheal Gillet","daniela Smith-jones")
str_to_title(my_string)
str_length(my_string)
str_count(my_string,"j")
```

Honestly, this is the kind of thing you "ChatGPT", "Google" or "ask-a-data-analyst-friend" when you need it. At least you know these tools exist: when you can **logically** explain in your mind what you want to do, it's doable.

---

# Dates and times

```{r}
library(lubridate)
today()
now()+60
today()+1
```

---

```{r}
adate <- "2022-06-01"
ymd(adate)
lubridate::dmy("26-06-19")
adatetime <- "2001-10-10 02:30:30"
ymd_hms(adatetime)

as.numeric(ymd_hms(adatetime))

```

---

```{r}
(my_data <- tibble(year=c("2022","2023"),
       month=c("6","1"),
       day=c("14","22")))

my_data |>
  mutate(date_string = paste0(year,"-",month,"_",day),
         date=lubridate::as_date(date_string))
```

---

```{r}
to_plot <- tibble(date=c("2019-01-01","2019-03-01","2020-01-01","2020-06-01"),value=rnorm(4,10,3))
ggplot(to_plot,aes(x=date,y=value)) + geom_point()
```
---
```{r}
to_plot <- to_plot |> mutate(date=ymd(date))
ggplot(to_plot,aes(x=date,y=value)) + geom_point() + geom_line() +
  scale_x_date(breaks = as_date(c("2019-01-01","2019-03-01","2020-01-01","2020-06-01")),
               labels=c("Jan 2019","March 2019","Jan 2020","June 202")) +
  theme(axis.text.x = element_text(angle=45)) + scale_y_continuous(limits=c(0,30))
```

---

# A worked example

See Script Code Example November 8